---
title: "Compartmental models"
author: "Ravi Brenner"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

Notes and code following chapter 2 of Modeling Infectious Diseases in Humans and Animals by Matt J. Keeling and Pejman Rohani.

```{r}
library(tidyverse)
library(deSolve)
```

## SIR model without demography
Basic model

$\frac{dS}{dt} = -\beta SI $

$\frac{dI}{dt} = \beta SI  - \gamma I$

$\frac{dR}{dt} = \gamma I$

beta here is the contact rate between individuals

gamma is the recovery rate, and 1/gamma is the average infectious period

R0 is beta / gamma

Note that S + I + R must = 1
```{r}
sirmod <- function(t, state, params){
  # pull listed inputs into environment as names variables
  with(as.list(c(state,params)),{
    # define equations
      dS = -beta * S * I 
      dI = beta*S*I - gamma*I
      dR = gamma * I
      
      list(c(dS,dI,dR))
  })
}

times = seq(0,100, by=1) #100 days, in discrete units
state = c(S = 0.999, I = 0.001, R = 0) # initial state values
params = c(beta = 2, gamma = 1/2, N = 1) # parameters

sir_out <- ode(y = state, t = times, func = sirmod, parms = params) |> 
  as_tibble()
```

plot model
```{r}
sir_out |>
  ggplot(aes(x = time)) + 
  geom_line(aes(y = S)) + 
  geom_line(aes(y = I),color = "red") + 
  geom_line(aes(y = R),color = "blue")
```

Final epidemic size
```{r}
final_size = rootSolve::runsteady(y=state,times=c(0,100), func=sirmod, parms=params)
final_size$y

#alternatively...assuming time is long enough
sir_out[nrow(sir_out),]
```


## SIR model with demography
Add in a mortality rate $\mu$, with average lifespan equal to $1/\mu$.

Here, we are also assuming that $\mu$ is the population's birth rate (hence the positive $\mu$ in the dS equation)

$\frac{dS}{dt} = \mu -\beta SI  - \mu S$

$\frac{dI}{dt} = \beta SI  - \gamma I - \mu I$

$\frac{dR}{dt} = \gamma I - \mu R$

R0 here is beta / (gamma + mu)

Note S + I + R here must still be =1
```{r}
sir_demo_mod <- function(t, state, params){
  # pull listed inputs into environment as names variables
  with(as.list(c(state,params)),{
    # define equations
      dS = mu -(beta * S * I ) - mu*S
      dI = (beta*S*I) - gamma*I - mu * I
      dR = gamma * I - mu*R
      
      # dead = mu*S + mu*I + mu*R
      # born = mu*N
      
      list(c(dS,dI,dR))
  })
}

times = seq(0,100, by=1) #100 days, in discrete units
state = c(S = 0.999, I = 0.001, R = 0) # initial state values
params = c(beta = 2, gamma = 1/2, mu = 0.0001, N = 1) # parameters

sir_demo_out <- ode(y = state, t = times, func = sir_demo_mod, parms = params) |> 
  as_tibble()
```

plot model
```{r}
sir_demo_out |>
  ggplot(aes(x = time)) + 
  geom_line(aes(y = S)) + 
  geom_line(aes(y = I),color = "red") + 
  geom_line(aes(y = R),color = "blue")
```

What happens at equilibrium when dS = dI = dR = 0?

There is a disease free equilibrium when S = 1, I = R = 0.

In endemic equilibrium, we find S* = (gamma + mu) / beta = 1/R0

Can plug that in to find that $I^* = \frac{\mu}{\gamma + \mu}(1-\frac{1}{R_0}) = \frac{\mu}{\beta}(R_0 -1)$

(R* is just 1-S* -I* here)

This requires R0 > 1 for endemic equilibrium, otherwise the disease-free equilibrium will hold.

```{r}
final_size = rootSolve::runsteady(y=state,times=c(0,10000), func=sir_demo_mod, parms=params)
final_size$y

r0 <- params["beta"] / (params["gamma"] + params["mu"])
I_star = (params["mu"] / params["beta"]) * (r0-1)
r0
I_star
```


Average age of infection

$A \approx \frac{1}{\mu(R_0-1)}$ (from dS equation). This is basically the inverse of the average period spent in susceptible class, which is approximated by the inverse of the force of infection, $1/\beta I^*$, subbing in I* from above.

Can rewrite this as $R_0 -1 \approx L/A$ with L=lifespan = 1/mu

```{r}
age_infex <- 1/(params["mu"]*(r0-1))
age_infex / 365
```



## SI models - infection induced mortality

There are multiple possibilities here, and things start to get more complex.

Mainly, consider probability $\rho$ of an infected individual dying from infection before recovering or dying from other causes.

This changes the dI equation to:

$\frac{dI}{dt} = \beta SI  - (\gamma +\mu)I - \frac{\rho}{1-\rho}(\gamma + \mu) I$

With the new term thought of as the per capita disease induced morality rate. We can rearrange this equation a bit:

$\frac{dI}{dt} = \beta SI   - \frac{\gamma + \mu}{1-\rho} I$

Note that as $\rho$ goes to 1, R0 drops to 0 (because new infections die almost immediately).

### Density dependent transmission
Up to now, we have been assuming S, I, and R are proportions and N = 1. Bus as N changes, the dynamics change too.

Note that we can no longer assume population size is fixed. A solution to this is to incorporate a fixed birth rate $\nu$ into the dS equation.

S->X
I->Y
R->Z

$\frac{dX}{dt} = \nu -\beta XY  - \mu X$

$\frac{dY}{dt} = \beta XY   - \frac{\gamma + \mu}{1-\rho} Y$

$\frac{dZ}{dt} = \gamma Y - \mu Z$


But since N can now vary, neet to consider $\beta XY$ in more detail.

$\frac{dN}{dt} = \nu - \mu N$, setting = 0 if no disease then $N = \nu/\mu$, the carrying capacity for the population. (disease free equilibrium solution for S*)

Now, taking into account a correction for reduced infectivity due to disease induced mortality, and a tern for the population size, $R_0 = \frac{\beta(1-\rho)\nu}{(\mu + \gamma)\mu}$

note now the endemic equilibrium:

$X^* = \frac{\mu + \gamma}{\beta(1-\rho)} = \frac{\nu}{\mu R_0}$

$Y^* = \frac{\mu}{\beta}(R_0-1)$

$Z^* = \frac{\gamma}{\beta}(R_0-1)$

$N^* = \frac{\nu}{\mu R_0}[1+(1-\rho)(R_0-1)]$

(Skipping over the frequency dependent version here, for simplicity). The takeaway here is as follows:

When disease-induced mortality is added to the SIR model with density-dependent transmission, the equilibrium and stability properties simply reflect a change in parameters.

When disease-induced mortality is added to the SIR model with frequency dependent transmission, the equilibrium and stability properties can change substantially, especially if the probability of mortality is high.

### Mortality late in infection
$\frac{dS}{dt} = \nu -\beta SI  -\mu S$

$\frac{dI}{dt} = \beta SI  - (\gamma+\mu) I$

$\frac{dR}{dt} = (1-\rho)\gamma I - \mu R$

Here, rho is the probability of an infected individual dying from the disease. Basically, they spend the entive infectious period in the I class, and then instead of moving to recovery, they are just removed from the R class as a death.

## SIS model (no immunity)

$\frac{dS}{dt} = \gamma I -\beta SI $

$\frac{dI}{dt} = \beta SI  - \gamma I $

S + I = 1

Can actually solve these to find that 

$I^* = (1-1/R_0)$

$S^* = 1/R_0$

$R_0 = \beta/\gamma$

As the equilibrium. Basically this means that for a disease that does not lead to long-term immunity, the disease will persis long term in the population if R0 > 1. Basically it will become endemic at a level determined by R0.

```{r}
sis_mod <- function(t, state, params){
  # pull listed inputs into environment as names variables
  with(as.list(c(state,params)),{
    # define equations
      dS = -(beta * S * I ) +gamma*I
      dI = (beta*S*I) - gamma*I 

      
      list(c(dS,dI))
  })
}

times = seq(0,100, by=1) #100 days, in discrete units
state = c(S = 0.999, I = 0.001) # initial state values
params = c(beta = 2, gamma = 1/2, N = 1) # parameters

sis_mod_out <- ode(y = state, t = times, func = sis_mod, parms = params) |> 
  as_tibble()

sis_mod_out |>
  ggplot(aes(x = time)) + 
  geom_line(aes(y = S)) + 
  geom_line(aes(y = I),color = "red")
```

## SIRS model - waning immunity

$\frac{dS}{dt} = \mu +\omega R -\beta SI  -\mu S$

$\frac{dI}{dt} = \beta SI  - (\gamma+\mu) I$

$\frac{dR}{dt} = \gamma I - \omega R -  \mu R$

Where $\omega$ is the rate at which immunity is lost and individuals go from R-->S. Similar to before, $R_0 = \beta /(\gamma + \mu)$


```{r}
sirs_mod <- function(t, state, params){
  # pull listed inputs into environment as names variables
  with(as.list(c(state,params)),{
    # define equations
      dS = mu -omega * R -(beta * S * I ) -mu*S
      dI = (beta*S*I) - (gamma+mu)*I 
      dR = gamma*I - omega*R - mu*R
      
      list(c(dS,dI,dR))
  })
}

times = seq(0,100, by=1) #100 days, in discrete units
state = c(S = 0.999, I = 0.001, R = 0) # initial state values
params = c(beta = 0.75, gamma = 1/2, N = 1, mu = 1/(70*12), omega = 1/(12*1)) # parameters

sirs_mod_out <- ode(y = state, t = times, func = sirs_mod, parms = params) |> 
  as_tibble()

sirs_mod_out |>
  ggplot(aes(x = time)) + 
  geom_line(aes(y = S)) + 
  geom_line(aes(y = I),color = "red") + 
  geom_line(aes(y = R),color = "blue")
```

## SEIR model - latent period

$\frac{dS}{dt} = \mu -\beta SI  -\mu S$

$\frac{dE}{dt} = \beta SI  -(\mu + \sigma) S$

$\frac{dI}{dt} = \sigma E - (\gamma+\mu) I$

$\frac{dR}{dt} = \gamma I -  \mu R$

The latent period (time spent infected but not infectious) is equal to 1/sigma.

And assume S + E + I + R = 1 (closed cohort).

This gives

$R_0 = \frac{\beta}{\mu+\gamma} \frac{\sigma}{\mu+\sigma}$

Basically, it's the same R0 as the SIR model, but some individuals in E die before reaching I, so they do not contribute to transmission. Usually, the second fraction is ~1 (since mu is very small compared to sigma--people's lifespan (in years) is usually longer than the time spend in latent period (often days or weeks)).

Although the SIR and SEIR models behave similarly at equilibrium (when the parameters are suitably rescaled), the SEIR model has a slower growth rate after pathogen invasion due to individuals needing to pass through the exposed class before they can contribute to the transmission process.


## carrier state

## discrete time models

## estimating parameters

## Putting it all together

S E I R C D N 
Births/deaths
Deaths late in infection


